Modeling memory and taxis to analyze migrations

### 2021-05-19 text/ms.Rmd
still working on a new model (!?!?!)

Diffusion-advection models have a long pedigree in animal movement modeling [@Skellam1951;@Turchin1998;@Okubo2001].  Essentially, these models are grounded in the general idea that animal movements - somewhat like movements of physical particles - might combine a random (diffusive) component with a directed (advective) component and that the two can be separated and modeled. While direct relationships between diffusion models and movement data are somewhat tenuous [@Gurarie2011a], as a theoretical tool for exploring processes they are invaluable for their versatility and the relative ease of numeric computation of the partial differential equations (PDEs) that are used to describe diffusion-advection models mathematically. Even as equations analytically intractable,

Thus, much theoretical and some applied work has been done on refining the basic assumptions of diffusion models, e.g. by including heterogeneity in populations [@Skalski2003;@Gurarie2009], fat-tailed dispersal kernels [@Kot1996], non-linear or otherwise complex responses to resources and consepecifics (**refs**).

However, perhaps the greatest difference between animals and the randomly moving passive particles that are best described by diffusion models is the cognitive abilities, including social behavior and memory. Refinements to diffusion-advection equations have revealed conditions under which non-local information gathering [@Fagan2017] and behavioral switching may confer foraging advantages [@Fagan2019], in particular when resources are dynamic and patchy. The interacting role of memory and sociality have been comparatively little studied.

Migration is a strategic behavior, typically involving a major trade-off between the costs of the migration against the benefits of accessing resources (whether food or predator refuge) that are highly seasonal and localized (**refs**).  While some migrants - both terrestrial and avian - follow a "green wave" of spring vegetation as it flowers across altitudinal or latitudinal gradients [@Bischof2012;@Kolzsch2015], many others "jump" the green wave, anticipating that critical resources will be available at the time of arrival (**refs**).

We anticiapted that under many conditons a blending of *tactical* (i.e. direct response to resource availability or perception) and *strategic* (i.e. memory-driven and forward-thinking) behavior can help a forager navigate a dynamic and heterogeneous resource environment, with a particular emphasis on the aquisition and adaptabilty of **migratory behavior** in a dynamic, seasonal resource environment To explore the interaction of memory and resource following as an adaptive strategy in dynamic environments. Specifically, we explore a one-dimensional advection diffusion model with long-term and short-term memory, and run the model in a dynamic, seasonal environment.

@@

where the variables $u(x,t)$ represents a population distributed in time and space, $h(x,t)$ is the distribution of the resource, $m(x,t,y)$ is the memory component in space and time at (discrete) year $y$, $\varepsilon$ is a rate of randomness (diffusion), $\alpha$ is the strength of advection in the direction of the resource $h(x,t)$ gradient), and $\beta$ is the strength of advection up the gradient memory.  The memory component we model as a mixture of a reference (long-term) memory, modified by more recent experiences as follows:
$$m(x,t) = \kappa^y \, u(x, t - y\tau) + \left(1-\kappa^y\right)\,u(x, t-\tau)\,\,\,\,(2)$$
Here, $y \in \{0,1,2,...\}$ represents the count of a given year, i.e. $y = \lfloor t/\tau \rfloor$. Thus, $u(x, t - y\tau)$ is the initial (year 0) "learned" collective behavior, $u(x, t-\tau)$ is the preceding year's collective behavior and $\kappa \in (0,1)$ is the relative contribution of the long-term (reference) memory.   When $\kappa = 1$, all of the memory is reference memory, and the system essentially responds to it's initial learned behavior as if it were a resource. If $\kappa = 0$, the memory is entirely short-term, and the process responds only to the behavior from the previous year.  With $0 < \kappa < 1$, the reference memory fades with time.  The model is confined to a one-dimensional bounded domain $[0,\chi]$, with no flux outside of the boundaries i.e. ${\frac{\partial u(0,t)}{\partial t}} = {\frac{\partial u(\chi,t)}{\partial t}} =0$,  Thus the total population is fixed at all times and found within the domain.

A few special cases of this model have been studied in the past. @Shigesada1979 (section 3) thoroughly analyzed the resource following diffusion model with no memory and a static resource (i.e.~$\beta = 0$ and $h(x,t) = h(x)$), showing that at equilibrium, the log of the population density is proportional to $h(x)/\alpha$.  @Shi2019 consider a single period memory model ($\kappa = 0$) in detail, with the addition of an independent birth-death process. Importantly, the authors prove the wellposedness (existence, uniqueness, and positivity) of this model, with arguments that are equally valid for our model. @Fagan2017 numerically explored an extension of this model with no memory but a dynamic resource and the addition of non-local information, i.e.~a model where the resource $h(x,t)$ is "perceived" via a convolution with a detection function, while @Fagan2019 added a behavioral switching behavior to the non-local information model.  This study

@@

We ran our numerical experiments on a spatial domain ranging from 0 to $\chi = 100$, and a period of duration $\tau = 100$ (i.e.~a 100 day "year").  We are interested in an approximately periodic resource dynamic, i.e. one in which $h(x,t \approx h(x, t-\tau))$ and $\tau$ is the period of the process. to do this.  Furthermore, we were mainly interested in a "non-surfable" resource where the patches are isolated and pulse periodically.

@@

Simulations are run until a (quasi)-equilibrium state is achieved, i.e. where the Bhattacharya index across subseqyuent years reaches a value of 0.9999.  Once that state is attained, we compute the metrics. We additionally collect the number of runs it

@@

>>>>

Seasonal migrations are an extremely widespread among terrestrial, aquatic, avian and invertebrate species [@Dingle2014].  For many species, migration is an extremely successful strategy, allowing a far greater number of individuals to inhabit landscape which might not otherwise be able to support large numbers year round [@Fryxell1996]. The ultimate evolutionary cause of seasonal migration, essentially, is that the energetic and survival related costs of migration are outweighed by the fitness benefits of accessing suitable seasonal resources, whether those are for energetic gain, predator avoidance, a suitable physical, biotic or social environment for reproduction [@Avgar2013].

Much as the ultimate causes of migration vary across species, the proximate causes, drivers and mechanism can vary considerably across and even within species [@Shaw2016]. Some migrants follow a "green wave" of spring vegetation as it flowers across altitudinal or latitudinal gradients [@Bischof2012;@Kolzsch2015]. These migrations can be considered "tactical", as they can occur - as an extreme simplification - purely as  response to local conditions.   Other migrants perform long-distance migrations in anticipation that critical resources will be available at the time of arrival at the end point of migration (**refs**).  This second behavior involves the greatest trade-off between the costs of migration against the benefits of accessing resources, whether food, suitable habitat for breeding, or predator refuge, that are highly seasonal and localized (**refs**).  This approach can be considered "strategic" in the sense that it is driven not by immediate cues but by a remembered anticipation [@Bracis2017].

Migratory species are often considered to be more vulnerable to environmental change, as a disruption in either of the seasonal ranges or along a migratory corridor can have significant negative impacts [@Wilcove2008;@Kauffman2021].  On the other hand, it has been argued that migratory species might be more resilient to disruptions due to their wide-ranging mobility [@Robinson2009]. Clearly, the ability to perform a migration without local cues in only possible if the behavior is hard-cued or remembered.  On the other hand, a strictly programmed behavior can be maladaptive if conditions change.

Here, we explore the relationship between memory and resource following for a migratory population under various static and dynamic seasonal resource distributions using a diffusion-advection model.

Diffusion-advection models have a long pedigree in animal movement modeling [@Skellam1951;@Turchin1998;@Okubo2001]. These models are grounded in the general idea that animal movements - somewhat like movements of physical particles - combine a random (diffusive) component with a directed (advective) component. While direct relationships between diffusion models and movement data are somewhat tenuous [@Gurarie2011a], as a theoretical tool for exploring processes they are invaluable for their versatility and the relative ease of numeric computation of the partial differential equations (PDEs) that are used to describe diffusion-advection models mathematically. Thus, much theoretical and some applied work has been done on refining the basic assumptions of diffusion models, e.g. by including heterogeneity in populations [@Skalski2003;@Gurarie2009], fat-tailed dispersal kernels [@Kot1996], non-linear or otherwise complex responses to resources and consepecifics (**refs**).

Perhaps the greatest difference between animals and randomly moving passive particles described by diffusion models is cognition, including social behavior and memory. Refinements to diffusion-advection equations have revealed conditions under which non-local information gathering [@Fagan2017] and behavioral switching may confer foraging advantages [@Fagan2019], in particular when resources are dynamic and patchy. The interacting role of memory and sociality, in contrast, have been comparatively little studied.

We anticipate that under many conditions a blending of *tactical* (i.e. direct response to resource availability or perception) and *strategic* (i.e. memory-driven and forward-thinking) behavior help foragers navigate dynamic, seasonal resource environments.

@@

where the variables $u(x,t)$ represents a population distributed in time and space, $h(x,t)$ is the distribution of the resource, $m(x,t,y)$ is the memory component in space and time at (discrete) year $y$, $\varepsilon$ is a rate of randomness (diffusion), $\alpha$ is the strength of advection in the direction of the resource $h(x,t)$ gradient), and $\beta$ is the strength of advection up the memory gradient.  The memory component we model as a mixture of a reference (long-term) memory, modified by more recent experiences as follows:
$$m(x,t) = \kappa^y \, u(x, t - y\tau) + \left(1-\kappa^y\right)\,u(x, t-\tau)\,\,\,\,(2)$$
Here, $y \in \{0,1,2,...\}$ represents the count of a given year, i.e. $y = \lfloor t/\tau \rfloor$. Thus, $u(x, t - y\tau)$ is the initial (year 0) "learned" collective behavior, $u(x, t-\tau)$ is the preceding year's collective behavior and $\kappa \in (0,1)$ is the relative contribution of the long-term (reference) memory.   When $\kappa = 1$, all of the memory is reference memory, and the system essentially responds to it's initial learned behavior as if it were a resource. If $\kappa = 0$, the memory is entirely short-term, and the process responds only to the behavior from the previous year.  With $0 < \kappa < 1$, the reference memory fades with time.  The model is confined to a one-dimensional bounded domain $[0,\chi]$, with no flux outside of the boundaries i.e. ${\frac{\partial u(0,t)}{\partial t}} = {\frac{\partial u(\chi,t)}{\partial t}} =0$,  Thus the total population is fixed at all times and found within the domain.

A few special cases of this model have been studied in the past. @Shigesada1979 (section 3) thoroughly analyzed the resource following diffusion model with no memory and a static resource (i.e.~$\beta = 0$ and $h(x,t) = h(x)$), showing that at equilibrium, the log of the population density is proportional to $h(x)/\alpha$.  @Shi2019 consider a single period memory model ($\kappa = 0$) in detail, with the addition of an independent birth-death process. Importantly, the authors prove the wellposedness (existence, uniqueness, and positivity) of this model, with arguments that are equally valid for our model. @Fagan2017 numerically explored an extension of this model with no memory but a dynamic resource and the addition of non-local information, i.e.~a model where the resource $h(x,t)$ is "perceived" via a convolution with a detection function. @Fagan2019 added a behavioral switching behavior to the non-local information model.  This study

Note, this model does not contain any explicit birth or death processes, the population is fixed, and the parameters remain constant throughout time.  Our interest is entirely in the ability of a fixed set of memory and movement parameters to ``navigate'' a dynamic environment.

@@

We ran our numerical experiments on a spatial domain ranging from 0 to $\chi = 100$, and a period of duration $\tau = 100$ (i.e.~a 100 day "year").  We are interested in an approximately periodic resource dynamic, i.e. one in which $h(x,t \approx h(x, t-\tau))$ and $\tau$ is the period of the process. Furthermore, we were mainly interested in a "non-surfable" resource where the patches are isolated and pulse periodically.

@@

Simulations are run until a (quasi)-equilibrium state is achieved, i.e. where the Bhattacharya index across subseqyuent years reaches a value of 0.9999.  Once that state is attained, we compute the mirgatoriness and foraging efficiency metrics, and additionally collect the number of years that were required to reach stationarity.

@@

Our model is somewhat similar to an evolutionary individual-based model [@Anderson2013], which showed that some inherent phenotypic plasticity is required to hedge both against persistent trends and increased stochasticity of resource distributions, but at the risk of sub-optimal access to resources [@Anderson2013].
